---
title: "R tutorial - R markdown"
author: "Wooyong Lee"
header-includes:
- \usepackage{amsfonts,amssymb,amsmath}
output:
  html_document
---
<!-- customize the title and the author in the above code -->

This is a tutorial for R markdown. It also serves as a sample R markdown code. Please check the source code of this document.

# What is R markdown?

- Think of it as a source code for creating a webpage that can embed R code

- You write a source code and "compile" it to create a webpage

# Creating an R markdown document

In RStudio, click **File -> New File -> R Markdown...**.

RStudio will create a skeleton file that contains basic explanations about R markdown. Read those.

You can compile R markdown file into HTML or PDF. I recommend HTML because it can adjust fontsize, linebreak, etc. according to the device.

# R markdown cheatsheet

- Check the R markdown cheatsheet:
https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf

- Especially, 
check **3. Markdown**.

- It covers most of 
what you need to know to write plain text

- Also check *Help -> Cheatsheets -> R Markdown Cheat Sheet* from RStudio

<!-- as in the cheatsheet, '*' also works for the unordered list -->

# LaTeX

- You can use LaTeX in R markdown.

- R markdown understands LaTeX syntax (at least those in the math mode) --- so just write them.

<!-- if you want to use LaTeX packages in R markdown, include them in the "header-includes" declaration in the header as this document did -->

# Writing math symbols and equations

Please check the source code of this document to see how the LaTeX code works in R markdown.

### Examples - read the source code!

<!-- $$ and $$ display a math equation in a separate line -->

Pythagorean theorem.

$$
  a^2 + b^2 = c^2
$$

<!-- $...$ creates an inline math expression -->

Arithmetic mean $\geq$ Geometric mean $\geq$ Harmonic mean.

$$
  \frac{1}{n}\sum_{i=1}^N X_i \geq \left( \prod_{i=1}^N X_i \right)^{1/N} \geq \frac{1}{\sum_{i=1}^N 1/X_i}.
$$

Greek letters, plain texts and multiple lines.

$$
  \begin{aligned}
  line 1 &= \alpha + \beta   \\
  \text{plain text} &= \gamma + \delta  \\
  line 3 &= \lambda + \sigma \\
  \textbf{bold plain text} &= \Lambda + \Sigma
  \end{aligned}
$$

Caligraphic, blackboard bold and bold.

$$
  \mathcal{F}, \mathbb{F}, \mathbf{F}.
$$

If you want to draw a math symbol but you don't know what to type, visit http://detexify.kirelabs.org/classify.html and draw the symbol using your mouse. For example, draw $\pm$ and check that the command is "pm".

# Embedding R

R markdown allows to run an R code within a document.

This allows us to alternate between plain text and R code.

This is handy for homework submission because you don't need to separately submit R code or copy R output into the document.

Below is an example of an R code (which we shall call "chunk") run within the document.

```{r}
Sys.setenv(lang="EN") ### this command sets the R locale to English ###
courseNo = "ECON 12345"
print(courseNo)
```

Note that it displays both code and result.

R chunks are cumulative. That is, the variable `courseNo` defined in the above can be used in later chunks:

```{r}
courseTitle = "Principles of Economics"
print(paste(courseNo, courseTitle, sep = " - "))
```

R markdown offers options that configure how R chunks are displayed and executed. For example, you can decide whether to hide the code, whether to hide the result, whether to display the error message, etc.

For a list of such options, check **5. Embed code** of the R markdown cheatsheet.

As an example, we will take a look at the `eval=FALSE` option, which tells R to display but not execute the code:

```{r, eval=FALSE}
SQRT2 = 1.414
```

In the above chunk, I specified `eval=FALSE`, meaning that the above chunk is not executed, which means `SQRT2` is not defined.

The below chunk confirms that `SQRT2` is not defined.

```{r, error=TRUE}
print(SQRT2)
```

If you check the source code of this document, you will notice that I specified `error=TRUE` in the above chunk. 

You must specify this option if you want R to print the error message. Otherwise the source code will not be compiled because an R chunk produced an error.






















