---
title: "R tutorial - basics"
author: "Wooyong Lee"
header-includes:
- \usepackage{amsfonts,amssymb,amsmath}
- \usepackage{graphicx}
- \usepackage{setspace}
- \usepackage{cleveref}
output:
  html_document:
    toc: false
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r header, echo=FALSE, message=FALSE, warning=FALSE}
Sys.setenv(lang="EN")
```

# basics

## value assignment

Some languages like C require to specify the type of the variable first and then assign numbers, e.g. `int abc; abc = 2;`.

But R does not require such thing. R changes the variable type automatically according to the variable type assigned.

```{r value assignment}
# the basic value assignment operator that always works
abc <- 2 # abc is set to a numeric variable
abc

# an easy-to-use operator that rarely, but not never, produces error
abc = 2
abc

# arrays
abc = c(1,2,3,4,5)
abc

# integer sequence
abc = 1:5
abc

# another integer sequence
abc = 3:(-2)
abc

# repeating vectors
abc = rep(2, 5)
abc

abc = rep(c(1,2,3), 5)
abc

# equispaced sequences
abc = seq(from=0, to=1, by=0.1)
abc

abc = seq(from=0, to=1, length=11)
abc

# matrices
abc = matrix(0, nrow=3, ncol=2)
abc

abc = matrix(1:6, nrow=3, ncol=2)
abc

abc = matrix(1:6, nrow=3, ncol=2, byrow=TRUE)
abc

abc = matrix(1:3, nrow=3, ncol=2)
abc

# multidimensional arrays
abc = array(0, dim=c(3,2,4))
abc

```

## arithmetic operators

Arithmetic operators are applied element-wise.

```{r arithmeric operators}
abc = matrix(1:6,  nrow=3, ncol=2)
def = matrix(7:12, nrow=3, ncol=2)
ghi = matrix(1:4,  nrow=2, ncol=2)

abc + def

abc * def

- def

# mathematical matrix multiplication
abc %*% ghi
```

## accessing matrices and arrays

```{r matrix, error=TRUE}
mat = matrix(1:6, nrow=3, ncol=2)
mat

# entries. The first row/column is indexed as 1.
mat[1,1]
mat[3,2]
mat[0,1] # there is no zero index in R.

# rows and columns
mat[3,]
mat[,2]

# submatrices
mat[c(1,2),]

# a useful trick
mat[c(1,3,3,3,2,1,3),]
```

## string handlers

```{r string}
abc = "X"
def = "Y"
paste(abc, def)
paste(abc, def, sep = "+")
paste(abc, def, sep = "")

paste(abc, 1:5, sep = "")

abc = 1:5
abc
paste(abc, collapse = "+")
```

## custom functions

The functions do not need to be written in separate files while some languages like MATLAB require to do so.

```{r custom functions}
# to write a custom function, assign FUNCTION to a VARIABLE
add = function(x,y) {
  return(x+y)
}
square = function(x) {
  return(x^2)
}

# the return statement can be omitted. R returns what is written in the last line.
square_noReturnStatement = function(x) {
  x^2
}

# to call a function, write the name and put arguments inside the bracket
abc = 3
add(x=abc, y=2)
square(x=3)
square_noReturnStatement(x=3)

# you can omit the labels
add(abc,2)

# you can assign the returned value to a variable
result = square(abc)
result
```

The following example explains what we mean by R automatically changing the variable types:

```{r auto assignment of variable types, error=TRUE}
# right now, a function is assigned to the variable "add":
add

# we know it works as a function
add(abc, 2)

# let's assign a number to "add"
add = 3;
add

# then the variable "add" is no longer a function
add(abc, 2)

```

## data.frame

`data.frame` is a variable type that handles datasets.

```{r data frame}
ctrs = c("U.S.", "Canada", "Mexico")
pops = c(318.9, 35.16, 122.3)

# you create a data frame as follows:
df = data.frame(Country=ctrs, Population=pops)
df
```

`data.frame` is a matrix with data manipulation capabilities.

```{r data frame 2}
# you can access the entries like a matrix
df[3,2]

# you can also access a column by its name. Use money symbol.
df$Population

# data.frame-specific functions are available, such as "subset"
subDF = subset(df, Population >= 100)
subDF

subDF = subset(df, Population >= 100, select = c("Country"))
subDF
```

## data.table

`data.table` is an enhancement of `data.frame` which is created by contributors of the R community.

`data.table` is not available with base R. To use it, we need to install a package, which is basically a collection of the codes. 

We install the package in our computers by the following code (requires internet connection to the R online database known as CRAN):

```{r datatableInstall, eval=FALSE}
# the name of the package that allows us to use data.table is "data.table".
install.packages("data.table")
```

Then we load the package by the following code:

```{r datatableLoad}
library(data.table)
```

`data.table` works like `data.frame`:

```{r datatable}
# recall:
ctrs
pops

# create datatable
dt = data.table(Country=ctrs, Population=pops)
dt
```

We can transform `data.frame` into `data.table`:

```{r datatable2}
# recall:
df
class(df)

# transform.
dt = as.data.table(df)
dt
class(dt)

```

`data.table` has additional features compared to `data.frame`. For example, to subset a data.table, we can do:

```{r datatable3}
# subset rows with pop >= 100
dt[Population >= 100]
# subset rows and choose the country and the population columns
dt[Population >= 100, .(Country)]
```

For more information, look at <https://github.com/Rdatatable/data.table/wiki/Getting-started>, which is a webpage created by the author of the `data.table` package.

## list

`list` is a variable type that can store other variables (sub-variables) in it. It is similar to `struct` in C.

```{r list}
abc = list(a=3, b="String", c=123.456)

# to access the sub-variables, call them using the money symbol:
abc$a
abc$b
abc$c
```

`list` can store anything as a sub-variable: functions, matrices, list inside a list, and so on.

The following example stores various model primitives of a matching model in labor economics.

```{r list2}
# model primitives
model = list(
  r = 0.05, # interest rate
  s = 0.02, # separation rate
  matchFunc = function(u,v) { 0.4 * u^0.5 * v^0.5 }, # matching function
  b = 0, # unemployment benefit
  c = 0, # cost of vacancy
  alpha = 0.5, # bargaining power
  prodFunc = function(x,y) { (x^(-2) + y^(-2))^(1/(-2)) + 0.2 } # production function
)
```

We can call a function inside the list in a usual way:

```{r list3}
model$prodFunc(1,2)
```

Instead of assigning names to the sub-variables, we can leave the names blank, in which case the serial numbers are assigned.

```{r list4}
abc = list(3, "String", 123.456)
abc

# to access the elements of a list, use the double brackets:
abc[[1]]
abc[[2]]
abc[[3]]
```

## syntax

The syntax of R is very similar to those of other languages. For example, the `for` loop looks like this:

```{r for loop}
# the canonical for loop with natural numbers
sum = 0
for(i in 1:4) {
  sum = sum + i
}
sum

# the loop works for ANY array
greekLetterArray = c("Alpha", "Beta", "Gamma", "Delta")

for(greekLetter in greekLetterArray) {
  print(greekLetter)
}

```

# ggplot2 package

## basics

`ggplot2` package is a package for plotting, as its name suggests.

To introduce various plots of `ggplot2`, let's use the `mtcars` dataset in the `datasets` package, which is preinstalled with base R.

```{r ggplot2 data}
library(datasets)
print(as.data.table(mtcars))
```

To install `ggplot2`:

```{r ggplot2 install, eval=FALSE}
install.packages("ggplot2")
```

Let's start from the scatterplot.

```{r ggplot2 scatterplot}
library(ggplot2)
ggplot(data=mtcars) + geom_point(aes(x=wt,y=mpg))
```

The syntax of `ggplot2` is summarized as the following:

- first, initialize the plot by `ggplot(data=mtcars)`.

- then, specify plotting options with a `+`.

In the above, `geom_point` specifies that we want to draw a scatterplot.

The `aes` argument specifies the variables that represent the x axis and the y axis. `aes` is an abbreviation for aesthetics.

The following example also specifies labels of the x and the y axes.

```{r ggplot2 scatterplot2}
ggplot(data=mtcars) + geom_point(aes(x=wt,y=mpg)) + xlab("Weight(lb)") + ylab("Miles per gallon")
```

`ggplot2` package allows you to save configurations.

```{r ggplot2 scatterplot3}
gg = ggplot(data=mtcars) + geom_point(aes(x=wt,y=mpg))
gg

gg + xlab("Weight(lb)") + ylab("Miles per gallon")
```

You can also specify the aesthetics when you initialize `ggplot`.

```{r ggplot2 scatterplot4}
gg = ggplot(data=mtcars, aes(x=wt,y=mpg))
gg + geom_point()
```

More options for aesthetics:

```{r ggplot2 scatterplot5}
# add colors according to the value of cyl
ggplot(data=mtcars) + geom_point(aes(x=wt,y=mpg,color=cyl))

# the function "factor" changes real numbers into categorical variables.
ggplot(data=mtcars) + geom_point(aes(x=wt,y=mpg,color=factor(cyl)))
```

The package's online documentation <http://docs.ggplot2.org/> has tons of examples of `ggplot2`.

For example, look at <http://docs.ggplot2.org/current/geom_point.html> for various usages of `geom_point`.

## theme

You may feel that the fonts and the legends are too small.

To change them, you need to modify the theme of the ggplot. 

You can do it for each plot by `+ theme(...)` (see online documentation <http://docs.ggplot2.org/current/theme.html>).

To permanently change the theme, use the `theme_update` function:

```{r ggplot2 theme}
# enlarges ggplot2 font and legend
theme_update(axis.text=element_text(colour="grey50",size=14),
             axis.title=element_text(size=14),
             strip.text=element_text(size=14), # these three lines are for larger axes labels and annotations
             legend.position="right", # this line specifies position of the legend
             legend.key.size=grid::unit(2.5,"lines"),
             legend.title=element_text(size=14,face="bold"),
             legend.text=element_text(size=14)) # these three lines are for larger legend texts

ggplot(data=mtcars) + geom_point(aes(x=wt,y=mpg,color=factor(cyl)))
```

## saving the plot

`ggsave` function saves the last displayed plot in a file.

You can specify the file extension, and DO NOT save in the bitmap image such as `.jpg` or `.bmp` since they do not scale up or down well.

Save your plot in a vector image such as `.pdf`.

```{r ggplot2 save, eval=FALSE}
# the basic command
ggsave("myplot.pdf")
# you can also specify the size
ggsave("myplot.pdf", width=8, height=5)
```

The file will be saved in your "working directory". Before saving the file, specify your working directory using the `setwd` function.

```{r setwd, eval=FALSE}
setwd("C:/GoogleDrive/MyWorkingDirectory/")
```

## more examples

We finish by looking at more `geom_xxx` functions. Google `geom_xxx ggplot2` and find the ggplot2 online documentation for more examples.

```{r ggplot2 geom_line}
ggplot(data=economics) + geom_line(aes(x=date, y=unemploy))

gg = ggplot(data=economics) + geom_line(aes(x=date, y=unemploy), size=1.2) 
gg + ggtitle("Number of unemployed in the U.S. (in thousands)")

gg + geom_smooth(aes(x=date,y=unemploy))
gg + geom_smooth(aes(x=date,y=unemploy), method="lm", se=FALSE)
```

```{r ggplot2 facet_grid}
gg = ggplot(data=mtcars) + geom_point(aes(x=wt,y=mpg,color=factor(cyl)))
gg

gg + facet_grid(. ~ cyl) 
```

```{r ggplot2 scale}
gg + scale_x_log10() + scale_y_log10()

gg + scale_x_continuous(trans = "log") + scale_y_continuous(trans = "log")
```

```{r ggplot2 density}
gg = ggplot(data=mtcars)

gg + geom_density(aes(x=mpg))

gg + geom_density(aes(x=mpg), size=1.2)

gg + geom_density(aes(x=mpg, color=factor(gear)), size=1.2)

gg + geom_density(aes(x=mpg, color=factor(gear), linetype=factor(gear)), size=1.2)

gg + geom_density(aes(x=mpg, color=factor(gear), linetype=factor(gear), fill=factor(gear)), size=1.2)

gg + geom_density(aes(x=mpg, color=factor(gear), linetype=factor(gear), fill=factor(gear)), size=1.2, alpha=0.2)
```

```{r ggplot2 geom_path}
gg = ggplot(data=economics) + geom_path(aes(x=unemploy/pop, y=psavert), size=1.2)
gg

# "geom_hline" does the same thing as "geom_abline" with slope=0.
gg + geom_abline(intercept = mean(economics$psavert), slope = 0) + geom_vline(xintercept = mean(economics$unemploy/(economics$pop)))

```




