---
title: "R tutorial - using C++ in R"
author: "Wooyong Lee"
header-includes:
- \usepackage{amsfonts,amssymb,amsmath}
- \usepackage{graphicx}
- \usepackage{setspace}
- \usepackage{cleveref}
output:
  html_document:
    toc: false
    toc_depth: 2
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r header, echo=FALSE, message=FALSE, warning=FALSE}
Sys.setenv(lang="EN")
library(microbenchmark) # this makes future library(microbenchmark) command display nothing.
library(Rcpp) # this makes future library(Rcpp) command display nothing.
```



# coding in C++ using Rcpp package

## boosting sequential loop

- We have seen that we can boost parallel loop using parallel computing.

- However, parallel computing is not useful for boosting sequential loop.

- One way to boost a sequential loop is to code it in faster langauges such as C++.

- However, learning and coding in C++ may require too much human time.

- We should spend most of the time in studying economics, not coding.



## Rcpp

- Rcpp package allows us to write a C++ code in R-like syntax.

- It also provides an interface between C++ and R.

- We only code the bottleneck in C++, and for the rest we use R.

- Still a little steep learning curve, but the gain can be huge.

- An Rcpp code can be more than ten times faster than the same R code.



## key differences between R and C++

In R:

- we execute the code line by line.

- we do not need to specify variables types. R automatically reads whether a variable is a number, a matrix, a data.frame, etc.

- we do not need to initialize variables. We can write `for(i in 1:3)` even if `i` was never defined before.

- array and matrix indices begin with `1`.

In C++:

- we write the entire code first, and then the compiler translates the entire code into a machine code.

- all variables must have a variable type, such as integer(`int`), real number(`double`), or string(`string`).

- all variables must be initialized (or **declared**) before use, even if it is a dummy variable in the for loop.

- array and matrix indices begin with `0`.

- every line must end with a `;`.



## C++ showcase

Consider the task of creating an array $(x, x^2, \ldots, x^k)$ given $x \in \mathbb{R}$ and $k \in \mathbb{N}$.

In R:

```{r loop R}

powers = function(x, k) {
  
  y = rep(0, k)
  
  for(j in 1:k) {
    y[j] = x^j
  }
  
  return(y)
}

```

Let's check its computation time:

```{r loop R benchmark}

x = 5
k = 1000
microbenchmark(
  r     = powers(x,k),
  times = 1000
)

```

Now we write the same code in C++.

- do not worry much about what these codes mean: we will use easier syntax of Rcpp

- just recognize that these are doing the same thing as the previous R code

```{Rcpp loop, eval=FALSE}
std::vector<double> powers_cpp(double x, int k) {
  
  std::vector<double> y(k);
  
  int j;
  
  for(j=0; j<k; j++) {
    y[j] = pow(x,j);
  }
  
  return y;
}
```

```{Rcpp loop Cpp, echo=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
std::vector<double> powers_cpp(double x, int k) {
  
  std::vector<double> y(k);
  
  int j;
  
  for(j=0; j<k; j++) {
    y[j] = pow(x,j);
  }
  
  return y;
}
```

We will discuss soon how to call this function in R.

Let's compare its performance to the R one:

```{r loop Cpp benchmark}
microbenchmark(
  r     = powers(x,k),
  cpp   = powers_cpp(x,k),
  times = 1000
)
```

We can see that C++ is much faster than R!

This verifies the common sense that a loop is very slow in R. 



## vectorization

If you are an experienced R coder, you should know that vectorization is very important for efficient computation in R.

In the above example, an experienced R coder should write 

```{r loop vectorR}

powers_vector = function(x, k) {
  
  return(x^(1:k))
  
}

```

We can see that its performance is between non-vectorized R and C++:

```{r loop vectorR benchmark}

microbenchmark(
  r     = powers(x,k),
  rvec  = powers_vector(x,k),
  cpp   = powers_cpp(x,k),
  times = 1000
)

```

This is because the vectorized code internally uses a loop that is more efficient than an R loop.



## coding like R in C++

We know C++ is much faster than R, but at the same time we want to use the easy syntax of R.

The Rcpp package provides a framework that allows us to use R-like syntax in C++.



## Rcpp variable types

Some of the key variable types defined in Rcpp:

- `NumericVector`: this corresponds to a 1-dim array in R, except that the length has to be pre-specified and fixed

- `NumericMatrix`: this corresponds to a matrix in R, except that the dimension has to be pre-specified and fixed

- `CharacterVector`: this stores character strings while `NumericVector` stores numbers

- `CharacterMatrix`: this stores character strings while `NumericMatrix` stores numbers

- `List`: this corresponds to a `list` in R.

In my experience, almost all of the R operations can be written in terms of these objects in C++.



## Rcpp "sugars"

The author of the Rcpp package created C++ versions of R functions that can be applied to the above variables.

These are called "sugars". A few examples:

- `dnorm`, `rnorm`, `pnorm`

- `abs`, `exp`, `sum`

- `match`, `max`, `min`

Google "Rcpp sugar" and you will find more functions.



## a simple Rcpp example

Consider the following R function that creates a matrix $Y = [X, X^2, X^3, ..., X^k]$ given a vector $X$ and an integer $k$:

```{r creatematrix R}

createMatrix = function(X, k) {
  
  # read dimension
  N = length(X)
  
  # prepare space for the matrix
  Y = matrix(0, nrow=N, ncol=k)
  
  # fill in the matrix column by column
  for(j in 1:k) {
    Y[,j] = X^j
  }
  
  # return matrix
  return(Y)
}

```

To write the same code in C++ and link it via Rcpp, we create a separate file `createMatrix.cpp` and write the following:

```{Rcpp creatematrix Cpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericMatrix createMatrix_cpp(NumericVector X, NumericVector k) {
  
  // read the "first element" of k
  int kk = k(0);
  
  // read length of X
  int N = X.size();
  
  // declare the Y matrix
  NumericMatrix Y(N,kk);
  
  // declare iterator used in the for loop
  int j;
  
  // fill in the matrix column by column
  for(j=0; j<kk; j++) {
    Y(_,j) = pow(X,j+1);
  }
  
  // return matrix
  return Y;
}
```

You can set the filename to be whatever you want, or you can write multiple functions in a single file. 

And then, we compile this code by the following command in R:

```{r createMatrix Cpp compile, eval=FALSE}
library(Rcpp)
sourceCpp("createMatrix.cpp")
```

This registers the function `createMatrix_cpp` in R environment, and we can call it just like an R function.

Now note the computational gain:

```{r creatematrix benchmark}

X = 1:100
k = 2000

microbenchmark(
  r     = createMatrix(x,k),
  cpp   = createMatrix_cpp(x,k),
  times = 1000
)

```

## the Rcpp code in depth

Recall the Rcpp code of the previous example:

```{Rcpp creatematrix Cpp recall, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericMatrix createMatrix_cpp(NumericVector X, NumericVector k) {
  
  // read the "first element" of k
  int kk = k(0);
  
  // read length of X
  int N = X.size();
  
  // declare the Y matrix
  NumericMatrix Y(N,kk);
  
  // declare iterator used in the for loop
  int j;
  
  // fill in the matrix column by column
  for(j=0; j<kk; j++) {
    Y(_,j) = pow(X,j+1);
  }
  
  // return matrix
  return Y;
}
```

Now we study this code in detail.

```{Rcpp, eval=FALSE}
#include <Rcpp.h>
using namespace Rcpp;
```

The first line is similar to package loading in R. It loads the Rcpp "package" by reading what is called the header file.

To understand the second line, note the following syntax in R:

- suppose we want to use the function `rq` (the quantile regression) in the package `quantreg`.

- in principle, we should type `quantreg::rq` to use the function `rq` in R.

- however, in R, we can omit `quantreg::` keyword. R automatically searches for all packages and finds the function `rq`.

- if multiple packages have the same function, we should specify the package name by the `::` keyword.

In C++, `namespace` corresponds to a package, and we access a namespace using the same `::` keyword.

The second line states that we will use objects and functions in Rcpp omitting the `Rcpp::` keyword.

```{Rcpp, eval=FALSE}
// [[Rcpp::export]]
NumericMatrix createMatrix_cpp(NumericVector X, NumericVector k) {
```

The first line is a keyword of Rcpp. When you compile the code, the Rcpp package will export all C++ functions to R that has the `// [[Rcpp::export]]` keyword above the function.

The second line specifies the function. Note that:

- we must specify the variable type of the return value, which is `NumericMatrix` here.

- we must specify the variable types of the arguments, which are `NumericVector`'s here.

```{Rcpp, eval=FALSE}
  // read the "first element" of k
  int kk = k(0);
```

This line performs variable type conversion. We need this because C++ is strict with variable types:

- `NumericVector` is a container of a numeric array, not the array itself.

- think of it as a list such as: `k = list(type="NumericVector", length=1, values=c(2000))`.

- what we need is `k$values[1]`, not `k` itself.

- Rcpp's `NumericVector` definition allows us to access values by `()` keyword.

- `k(0)` returns `k$values[1]`, and we save it as an integer variable `kk`.

In my experience, most of the coding error in Rcpp is due to the variable type conversion problems. Understanding how to match variable types is very important in Rcpp.

In practice, you can simply declare the function as `NumericMatrix createMatrix_cpp(NumericVector X, int k)`. When you call it from R with `k=3`, then `3`, which is a `NumericVector` with length 1, will be automatically converted to `int` in C++. In my example I made such conversion explicit to explain the concept.

```{Rcpp, eval=FALSE}
  // read length of X
  int N = X.size();
```

- In C++, objects often have functions inside it (compare it to R where we can write functions as a member of `List`).

- We can call such functions using `.` keyword.

- The `size` function of `NumericVector` returns the length of the vector.

```{Rcpp, eval=FALSE}
  // declare the Y matrix
  NumericMatrix Y(N,kk);
```

This line declares the Y matrix.

The code is self-explanatory: it creates a `NumericMatrix` variable `Y` with `N` rows and `kk` columns.

```{Rcpp, eval=FALSE}
  // declare iterator used in the for loop
  int j;
  
  // fill in the matrix column by column
  for(j=0; j<kk; j++) {
    Y(_,j) = pow(X,j+1);
  }
```

This implements the R for loop. Note that the code is similar to the R code. Especially, compare 

- `Y[,j] = X^j`

and

- `Y(_,j) = pow(X,j+1);`

Instead of declaring `j` first and then using it in the for loop, you can simply type `for(int j=0; j<kk; j++)`, which is a shorthand for the above code.

```{Rcpp, eval=FALSE}
  // return matrix
  return Y;
```

This returns the Y matrix as a function value. Unlike R, we cannot omit the `return` keyword.

## Rcpp cheatsheet

As is the case with other R packages, it has a cheatsheet:
http://dirk.eddelbuettel.com/code/rcpp/Rcpp-quickref.pdf

It explains various basic operations that can be done with NumericVector and NumericMatrix, with examples.



## an example - simulated likelihood

Now we showcase a more realistic example. Consider a panel data random effect model with $T$ periods where

$$
  Y_{it} = \beta_{i0} + \beta_{i1} Y_{i,t-1} + \varepsilon_{it}, \quad t=1, \ldots, T,
$$

where $(\beta_{i0}, \beta_{i1})$ is a random effect such that

$$
  \left[\begin{array}{c} \beta_{i0} \\ \beta_{i1} \end{array}\right] \sim
  N \left(
  \left[\begin{array}{c} b_{00} + b_{01}Y_{i0} \\ b_{10} + b_{11}Y_{i0} \end{array}\right],
  ~I_2
  \right),
$$

and $\varepsilon_{it}$ is a random variable independent of $(\beta_{i0},\beta_{i1},Y_{i0},\ldots,Y_{i,t-1})$ such that

$$
  \varepsilon_{it} \sim N(0, s_t),
$$

where $p \equiv (b_{00}, b_{01}, b_{10}, b_{11}, s_1, \ldots, s_T)$ are the parameters to be estimated.

Given the parameter value $p$, the log-likelihood of this model is given by

$$
  l(p) = \sum_{i=1}^N l_i(p) = \sum_{i=1}^N 
  \ln\left( 
  \int \prod_{t=1}^T \phi\left(\frac{Y_{it} - \beta_{i0} - \beta_{i1}Y_{i,t-1}}{s_t}\right) 
  f(\beta_{i0}, ~\beta_{i1}; p) d(\beta_{i0}, \beta_{i1})
  \right).
$$

where $f$ is the density of $(\beta_{i0}, \beta_{i1})$ given $p$ according to the specification. The simulated likelihood approach evaluates this likelihood by simulating a sample from $f$.

Our task is to compute the array $(l_1, \ldots, l_N)$ via simulated likeliood given

- the $N \times (T+1)$ matrix $Y = [Y_{it}]_{i=1, t=0}^{N, T}$, and 

- the $S \times 2$ matrix that contains $S$ standard bivariate normal draws $Z = [(Z_{s0}, Z_{s1})]_{s=1}^S$.

The individual likelihood array written in R is:

```{r simLik R}
indivLogLik = function(p, Y, Z) {
  
  # read dimensions
  N = nrow(Y)
  T = ncol(Y) - 1
  S = nrow(Z)
  
  # a vector that stores (l_1, ..., l_N)
  lik = rep(0, N)
  
  # for each i,
  for(i in 1:N) {
    
    # compute beta0 and beta1 values from the bivariate standard normal draws
    beta0 = p[1] + p[2] * Y[i,1] + Z[,1]
    beta1 = p[3] + p[4] * Y[i,1] + Z[,2]
    
    # initialize a vector that stores simulated likelihood values
    simulatedLik = rep(0, S)
    
    # now evaluate the likelihoods by looping over t
    for(t in 1:T) {
      # compute simulated likelihoods in logs
      simulatedLik = simulatedLik + dnorm(Y[i,t+1] - beta0 - beta1 * Y[i,t], mean=0, sd=p[4+t], log=TRUE)
    }
    
    # exponentiate and average, and then store the l_i value
    lik[i] = mean(exp(simulatedLik))
  }
  
  # return l_i
  return(log(lik))
}
```

Note that the double loop of $i$ and $t$ cannot be reduced by vectorization because the distribution of $(\beta_{i0}, \beta_{i1})$ differs across individuals.

The same code can be written in C++, only using NumericVector and NumericMatrix as key objects, as follows:

```{Rcpp simLik Cpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector indivLogLik_cpp(NumericVector p, NumericMatrix Y, NumericMatrix Z) {
  
  // read dimensions
  int N = Y.nrow();
  int T = Y.ncol()-1;
  int S = Z.nrow();
  
  // a vector that stores (l_1, ..., l_N)
  NumericVector lik(N);
  
  // In the R code, we used some auxiliary variables in the loop
  // In C++, we need to prepare them in advance
  NumericVector beta0(S), beta1(S), simulatedLik(S);
  
  // the same holds for the iterators
  int i;
  int t;
  
  // for each i,
  for(i=0; i<N; i++) {
    
    // compute beta0 and beta1 values from the bivariate standard normal draws
    // note that, in C++, the label begins with 0 and not 1
    beta0 = p(0) + p(1) * Y(i,0) + Z(_,0);
    beta1 = p(2) + p(3) * Y(i,0) + Z(_,1);
    
    // initialize a vector that stores simulated likelihood values
    simulatedLik = rep(0, S);
    
    for(t=0; t<T; t++) {
      // compute simulated likelihoods in logs
      simulatedLik = simulatedLik + dnorm(Y(i,t+1) - beta0 - beta1 * Y(i,t), 0, p(3+t), true);
    }
    
    // exponentiate and average, and then store the l_i value
    lik(i) = mean(exp(simulatedLik));
  }
  
  // return l_i
  return log(lik);
}
```

Now let's compare the performance:

```{r simLik benchmark}
# prepare some data
.N = 1000
.T = 10
.S = 1000

p = c(rnorm(4), exp(rnorm(.T)))
Y = matrix(rnorm(.N * (.T+1)), nrow=.N, ncol=.T+1)
Z = matrix(rnorm(.S * 2)     , nrow=.S, ncol=2)

# run benchmark
library(microbenchmark)
microbenchmark(
  R     = indivLogLik(p, Y, Z),
  Cpp   = indivLogLik_cpp(p, Y, Z),
  times = 3
)
```

There is not much computational gain here compared to other simple examples, but I hope this example shows that you can replace an R code with a C++ code in a R-like syntax using NumericVector and NumericMatrix.



## you don't need to write everything in C++

In practice, you might face a lot of variable type conversion problems in C++.

For example, in the simulated likelihood example, suppose you want the function to return $l$ instead of the $(l_1, \ldots, l_N)$ array. 

The first thing you would do is to replace the last line with

```{Rcpp wrongReturn, eval=FALSE}
return sum(log(lik));
```

However, this will incur a variable type conversion problem. 

This is because the `sum` function returns a `double` variable type, which is not a `NumericVector` type. 

One correct way is to write

```{Rcpp correctReturn, eval=FALSE}
NumericVector sumLik(1);
sumLik(0) = sum(log(lik));
return sumLik;
```

where we save the value `sum(log(lik))` in `sumLik`, which is a `NumericVector` array of length 1, and then return `sumLik`.

However, there is a better solution (in terms of the human time). You can just write a wrapper of the C++ function in R:

```{r wrapper}
logLik = function(p, Y, Z) {
  
  likelihoods = indivLogLik_cpp(p, Y, Z)
  return(sum(likelihoods))
  
}
```

You don't need to modify your C++ function. Just get the array returned from C++ and then sum it up in R.

Remember to write only the bottleneck in C++ to gain performance and to write the rest in R to save human time! 



## creating a package with Rcpp

If you use Rcpp in your research, these two problems may become an issue:

- every time you run your code, you need to compile the C++ code using `sourceCpp` which may take a while

- it does not work well with parallel computing: the parallel backends cannot compile the C++ code

A recommended solution for these problems is to create a package that contains the C++ functions.

- When you install the package, your R session will compile your C++ code. After that, when you load the package, it will just load the complied code which is instant.

- You can load the package to the parallel backends and use parallel computing.

Thankfully, creating a package is easy with RStudio:

- click **File -> New Project... -> New Directory -> R package**.

- click **Type** to change it to **Package w/ Rcpp**.

- include your C++ source code and create project.

- create descriptions and documentations (you can create dummy documentations if you are the only user of your functions).

- build a package by clicking **clean and rebuild** first and then **create a source package**.

When I was using this trick by myself, I learned a lot from this document:
http://web.mit.edu/insong/www/pdf/rpackage_instructions.pdf

The document is written for beginners, so it would be easy to understand.







